A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-23 paper.Rmd
done with 5.1

Data collection used a total of `r filter(vi_survey, !attention_check) %>% count(unique_lineup_id) %>% nrow()` lineups, and resulted in a total of `r nrow(filter(vi_survey, !attention_check))` evaluations from `r vi_survey %>% count(set, exp) %>% nrow()` participants. Roughly half corresponded to the two models, non-linearity and heteroskedasticiy, and the three collection periods had similar numbers in each. Each participant received two of the `r vi_survey %>% filter(attention_check) %>% count(unique_lineup_id) %>% nrow()` attention check lineups which were used to filter results of participants who were clearly not making an honest effort (only 11 of `r vi_survey %>% count(set, exp) %>% nrow() + 11`). To estimate $\alpha$ for calculating statistical significance (see Section \ref{sig}) there were `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations of `r vi_survey %>% filter(null_lineup, !attention_check) %>% count(unique_lineup_id) %>% nrow()` null lineups. Neither the attention checks nor null lineups were used for the subsequent analysis. The de-identified data, `vi_survey`, is made available in the R package, `visage`.

The data was collected on lineups constructed from four different fitted value distributions: uniform, normal, skewed and discrete. More data was collected on the uniform (each evaluated by 11 participants) than the others (each evaluated by 5 participants). The analysis in Sections \ref{power-analysis}-\ref{hetero-analysis}, uses only results from lineups generated with uniform fitted values, a total `r nrow(filter(vi_survey, !attention_check, !null_lineup, x_dist == "uniform"))` lineup evaluations. This was decided in order to compare the conventional and visual test performance for an optimal scenario. Section \ref{effect-of-fitted-value-distributions} examines how the results may be affected if the fitted value distribution was different.

@@

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power curves of various tests plotted against the effect size in the residuals, for non-linearity and heteroskedasticity, respectively. In each case the power of visual test is calculated for multiple bootstrap samples leading to the many (orange) curves. The effect size was computed at a 5% significance level and plotted on a natural logarithmic scale. To facilitate visual calibration of effect size values with the corresponding diagnostic plots, a sequence of example residual plots with increasing effect sizes is provided at the bottom of these figures. These plots serve as a visual aid to help readers understand how different effect size values translate to changes in the diagnostic plots. The horizontal lines of dots at 0 and 1 represent the non-reject or reject made by visual tests for each lineup.

@@

Figure \ref{fig:polypower} compares the power for the different tests for non-linear structure in the residuals. The test with the uniformly higher power is the RESET test, one that specifically tests for non-linearity. (Note that the BP and SW tests have much lower power, which is expected because they are not designed to detect non-linearity.) The bootstrapped power curves for the visual test are effectively a shift right from that of the RESET test. This means that the RESET test will reject at a lower effect size (less structure) than the visual test, but otherwise the performance will be similar. In other words, the RESET test is more sensitive than the visual test. This is not necessarily a good feature for the purposes of diagnosing model defects. If one scans the residual plot examples at the bottom, we might argue that the non-linearity is not sufficiently problematic until an effect size of around 3 or 3.5. The RESET test would reject closer to an effect size of 2, but the visual test would reject closer 3.25, for a significance level of 0.05.

For the heteroskedasticity pattern, the power of BP test, designed for detecting heteroskedasticity, is uniformly higher than the other tests. The visual test power curve is a right shift. This shows a similar story to the power curves for non-linearity pattern - the conventional test is more sensitive than the visual test. From the example residual plots at the bottom we might argue that the heteroskedasticity becomes noticeably visible around an effect size of 3 or 3.5. However the BP test would reject at around effect size 2.5. Interestingly, the power curve for the SW test (for non-normality) is only slightly different to that of the visual test, suggesting that it performs reasonably for detecting heteroskedasticity, too. The power curve for the BP test suggests it is not useful for detecting heteroskeadsticity, as expected.

Overall, the results show that the conventional tests are more sensitive than the visual test. The conventional tests do have higher power for the patterns they are designed to detect, and are generally unable to detect other patterns. The visual test doesn't require specifying the pattern ahead of time, relying purely on whether the observed residual plot is detectably different from "good" residual plots. They will perform equally well regardless of the type of model defect. This aligns with the advice of experts on residual analysis, who consider residual plot analysis to be an indispensable tool for diagnosing model problems. What we gain from using a visual test for this purpose is the removal of any subjective arguments about whether a pattern is visible or not. The lineup protocol provides the calibration for detecting patterns: that if the pattern in the data plot cannot be distinguished from patterns in good residual plots, then no discernible problem with the model exists.

>>>>

Data collection used a total of `r filter(vi_survey, !attention_check) %>% count(unique_lineup_id) %>% nrow()` lineups, and resulted in a total of `r nrow(filter(vi_survey, !attention_check))` evaluations from `r vi_survey %>% count(set, exp) %>% nrow()` participants. Roughly half corresponded to the two models, non-linearity and heteroskedasticiy, and the three collection periods had similar numbers of evaluations. Each participant received two of the `r vi_survey %>% filter(attention_check) %>% count(unique_lineup_id) %>% nrow()` attention check lineups which were used to filter results of participants who were clearly not making an honest effort (only 11 of `r vi_survey %>% count(set, exp) %>% nrow() + 11`). To estimate $\alpha$ for calculating statistical significance (see Section \ref{sig}) there were `r nrow(filter(vi_survey, null_lineup, !attention_check))` evaluations of `r vi_survey %>% filter(null_lineup, !attention_check) %>% count(unique_lineup_id) %>% nrow()` null lineups. Neither the attention checks nor null lineups were used in the subsequent analysis. The de-identified data, `vi_survey`, is made available in the R package, `visage`.

The data was collected on lineups constructed from four different fitted value distributions: uniform, normal, skewed and discrete. More data was collected on the uniform distribution (each evaluated by 11 participants) than the others (each evaluated by 5 participants). The analysis in Sections \ref{power-analysis}-\ref{hetero-analysis} uses only results from lineups generated with uniform fitted values, for a total `r nrow(filter(vi_survey, !attention_check, !null_lineup, x_dist == "uniform"))` lineup evaluations. This allows us to compare the conventional and visual test performance in an optimal scenario. Section \ref{effect-of-fitted-value-distributions} examines how the results may be affected if the fitted value distribution was different.

@@

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power curves of various tests plotted against the effect size in the residuals for non-linearity and heteroskedasticity respectively. In each case the power of visual test is calculated for multiple bootstrap samples leading to the many (orange) curves. The effect size was computed at a 5% significance level and plotted on a natural logarithmic scale. To facilitate visual calibration of effect size values with the corresponding diagnostic plots, a sequence of example residual plots with increasing effect sizes is provided at the bottom of these figures. These plots serve as a visual aid to help readers understand how different effect size values translate to changes in the diagnostic plots. The horizontal lines of dots at 0 and 1 represent the non-rejection or rejection decisions made by visual tests for each lineup.

@@

Figure \ref{fig:polypower} compares the power for the different tests for non-linear structure in the residuals. The test with the uniformly higher power is the RESET test, one that specifically tests for non-linearity. Note that the BP and SW tests have much lower power, which is expected because they are not designed to detect non-linearity. The bootstrapped power curves for the visual test are effectively a shift right from that of the RESET test. This means that the RESET test will reject at a lower effect size (less structure) than the visual test, but otherwise the performance will be similar. In other words, the RESET test is more sensitive than the visual test. This is not necessarily a good feature for the purposes of diagnosing model defects: If we scan the residual plot examples at the bottom, we might argue that the non-linearity is not sufficiently problematic until an effect size of around 3 or 3.5. The RESET test would reject closer to an effect size of 2, but the visual test would reject closer 3.25, for a significance level of 0.05. The visual test matches the robustness of the model to (minor) violations of assumptions much better.

For the heteroskedasticity pattern, the power of BP test, designed for detecting heteroskedasticity, is uniformly higher than the other tests. The visual test power curve is a right shift. This shows a similar story to the power curves for non-linearity pattern: the conventional test is more sensitive than the visual test. From the example residual plots at the bottom we might argue that the heteroskedasticity becomes noticeably visible around an effect size of 3 or 3.5. However the BP test would reject at around effect size 2.5. Interestingly, the power curve for the SW test (for non-normality) is only slightly different to that of the visual test, suggesting that it performs reasonably for detecting heteroskedasticity, too. The power curve for the BP test suggests it is not useful for detecting heteroskeadsticity, as expected.

Overall, the results show that the conventional tests are more sensitive than the visual test. The conventional tests do have higher power for the patterns they are designed to detect, and are generally unable to detect other patterns. The visual test doesn't require specifying the pattern ahead of time, relying purely on whether the observed residual plot is detectably different from "good" residual plots. They will perform equally well regardless of the type of model defect. This aligns with the advice of experts on residual analysis, who consider residual plot analysis to be an indispensable tool for diagnosing model problems. What we gain from using a visual test for this purpose is the removal of any subjective arguments about whether a pattern is visible or not. The lineup protocol provides the calibration for detecting patterns: if the pattern in the data plot cannot be distinguished from the patterns in good residual plots, then no discernible problem with the model exists.

### 2023-04-23 paper.tex
done with 5.1

Data collection used a total of 1152 lineups, and resulted in a total of
7974 evaluations from 443 participants. Roughly half corresponded to the
two models, non-linearity and heteroskedasticiy, and the three
collection periods had similar numbers in each. Each participant
received two of the 24 attention check lineups which were used to filter
results of participants who were clearly not making an honest effort
(only 11 of 454). To estimate \(\alpha\) for calculating statistical
significance (see Section \ref{sig}) there were 720 evaluations of 36
null lineups. Neither the attention checks nor null lineups were used
for the subsequent analysis. The de-identified data,
\texttt{vi\_survey}, is made available in the R package,
\texttt{visage}.

The data was collected on lineups constructed from four different fitted
value distributions: uniform, normal, skewed and discrete. More data was
collected on the uniform (each evaluated by 11 participants) than the
others (each evaluated by 5 participants). The analysis in Sections
\ref{power-analysis}-\ref{hetero-analysis}, uses only results from
lineups generated with uniform fitted values, a total 3069 lineup
evaluations. This was decided in order to compare the conventional and
visual test performance for an optimal scenario. Section
\ref{effect-of-fitted-value-distributions} examines how the results may
be affected if the fitted value distribution was different.

@@

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the
residuals, for non-linearity and heteroskedasticity, respectively. In
each case the power of visual test is calculated for multiple bootstrap
samples leading to the many (orange) curves. The effect size was
computed at a 5\% significance level and plotted on a natural
logarithmic scale. To facilitate visual calibration of effect size
values with the corresponding diagnostic plots, a sequence of example
residual plots with increasing effect sizes is provided at the bottom of
these figures. These plots serve as a visual aid to help readers
understand how different effect size values translate to changes in the
diagnostic plots. The horizontal lines of dots at 0 and 1 represent the
non-reject or reject made by visual tests for each lineup.

@@

Figure \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. (Note that the BP and SW tests have much lower power,
which is expected because they are not designed to detect
non-linearity.) The bootstrapped power curves for the visual test are
effectively a shift right from that of the RESET test. This means that
the RESET test will reject at a lower effect size (less structure) than
the visual test, but otherwise the performance will be similar. In other
words, the RESET test is more sensitive than the visual test. This is
not necessarily a good feature for the purposes of diagnosing model
defects. If one scans the residual plot examples at the bottom, we might
argue that the non-linearity is not sufficiently problematic until an
effect size of around 3 or 3.5. The RESET test would reject closer to an
effect size of 2, but the visual test would reject closer 3.25, for a
significance level of 0.05.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve is a right shift. This shows a similar story
to the power curves for non-linearity pattern - the conventional test is
more sensitive than the visual test. From the example residual plots at
the bottom we might argue that the heteroskedasticity becomes noticeably
visible around an effect size of 3 or 3.5. However the BP test would
reject at around effect size 2.5. Interestingly, the power curve for the
SW test (for non-normality) is only slightly different to that of the
visual test, suggesting that it performs reasonably for detecting
heteroskedasticity, too. The power curve for the BP test suggests it is
not useful for detecting heteroskeadsticity, as expected.

Overall, the results show that the conventional tests are more sensitive
than the visual test. The conventional tests do have higher power for
the patterns they are designed to detect, and are generally unable to
detect other patterns. The visual test doesn't require specifying the
pattern ahead of time, relying purely on whether the observed residual
plot is detectably different from ``good'' residual plots. They will
perform equally well regardless of the type of model defect. This aligns
with the advice of experts on residual analysis, who consider residual
plot analysis to be an indispensable tool for diagnosing model problems.
What we gain from using a visual test for this purpose is the removal of
any subjective arguments about whether a pattern is visible or not. The
lineup protocol provides the calibration for detecting patterns: that if
the pattern in the data plot cannot be distinguished from patterns in
good residual plots, then no discernible problem with the model exists.

>>>>

Data collection used a total of 1152 lineups, and resulted in a total of
7974 evaluations from 443 participants. Roughly half corresponded to the
two models, non-linearity and heteroskedasticiy, and the three
collection periods had similar numbers of evaluations. Each participant
received two of the 24 attention check lineups which were used to filter
results of participants who were clearly not making an honest effort
(only 11 of 454). To estimate \(\alpha\) for calculating statistical
significance (see Section \ref{sig}) there were 720 evaluations of 36
null lineups. Neither the attention checks nor null lineups were used in
the subsequent analysis. The de-identified data, \texttt{vi\_survey}, is
made available in the R package, \texttt{visage}.

The data was collected on lineups constructed from four different fitted
value distributions: uniform, normal, skewed and discrete. More data was
collected on the uniform distribution (each evaluated by 11
participants) than the others (each evaluated by 5 participants). The
analysis in Sections \ref{power-analysis}-\ref{hetero-analysis} uses
only results from lineups generated with uniform fitted values, for a
total 3069 lineup evaluations. This allows us to compare the
conventional and visual test performance in an optimal scenario. Section
\ref{effect-of-fitted-value-distributions} examines how the results may
be affected if the fitted value distribution was different.

@@

Figures \ref{fig:polypower} and \ref{fig:heterpower} present the power
curves of various tests plotted against the effect size in the residuals
for non-linearity and heteroskedasticity respectively. In each case the
power of visual test is calculated for multiple bootstrap samples
leading to the many (orange) curves. The effect size was computed at a
5\% significance level and plotted on a natural logarithmic scale. To
facilitate visual calibration of effect size values with the
corresponding diagnostic plots, a sequence of example residual plots
with increasing effect sizes is provided at the bottom of these figures.
These plots serve as a visual aid to help readers understand how
different effect size values translate to changes in the diagnostic
plots. The horizontal lines of dots at 0 and 1 represent the
non-rejection or rejection decisions made by visual tests for each
lineup.

@@

Figure \ref{fig:polypower} compares the power for the different tests
for non-linear structure in the residuals. The test with the uniformly
higher power is the RESET test, one that specifically tests for
non-linearity. Note that the BP and SW tests have much lower power,
which is expected because they are not designed to detect non-linearity.
The bootstrapped power curves for the visual test are effectively a
shift right from that of the RESET test. This means that the RESET test
will reject at a lower effect size (less structure) than the visual
test, but otherwise the performance will be similar. In other words, the
RESET test is more sensitive than the visual test. This is not
necessarily a good feature for the purposes of diagnosing model defects:
If we scan the residual plot examples at the bottom, we might argue that
the non-linearity is not sufficiently problematic until an effect size
of around 3 or 3.5. The RESET test would reject closer to an effect size
of 2, but the visual test would reject closer 3.25, for a significance
level of 0.05. The visual test matches the robustness of the model to
(minor) violations of assumptions much better.

For the heteroskedasticity pattern, the power of BP test, designed for
detecting heteroskedasticity, is uniformly higher than the other tests.
The visual test power curve is a right shift. This shows a similar story
to the power curves for non-linearity pattern: the conventional test is
more sensitive than the visual test. From the example residual plots at
the bottom we might argue that the heteroskedasticity becomes noticeably
visible around an effect size of 3 or 3.5. However the BP test would
reject at around effect size 2.5. Interestingly, the power curve for the
SW test (for non-normality) is only slightly different to that of the
visual test, suggesting that it performs reasonably for detecting
heteroskedasticity, too. The power curve for the BP test suggests it is
not useful for detecting heteroskeadsticity, as expected.

Overall, the results show that the conventional tests are more sensitive
than the visual test. The conventional tests do have higher power for
the patterns they are designed to detect, and are generally unable to
detect other patterns. The visual test doesn't require specifying the
pattern ahead of time, relying purely on whether the observed residual
plot is detectably different from ``good'' residual plots. They will
perform equally well regardless of the type of model defect. This aligns
with the advice of experts on residual analysis, who consider residual
plot analysis to be an indispensable tool for diagnosing model problems.
What we gain from using a visual test for this purpose is the removal of
any subjective arguments about whether a pattern is visible or not. The
lineup protocol provides the calibration for detecting patterns: if the
pattern in the data plot cannot be distinguished from the patterns in
good residual plots, then no discernible problem with the model exists.
